The role of aneuploidy in the evolution of cancer drug resistance

### 2023-06-21 EvolutionaryRescue1.tex
comments

To estimate the rescue probability $p_{rescue}$, we assume independence between clonal lineages starting from an initial population of $N$ wildtype cells (we check the effect of density-dependent growth on our results below).
Thus, the rescue probability is given by 
\begin{subequations}
\begin{align}
p_{rescue} &= 1-\left(1-p_w\right)^N \label{rescue_prob} \\
& \approx 1-\e^{-Np_w} , \label{rescue_prob2}
\end{align}
\end{subequations}
where the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not $N p_w$) is small.

Applying the approximations for the survival probability $p_w$ from \cref{survprobwapprox1,survprobwinitial,scenario3} in \cref{rescue_prob2} and substituting $\Delta_k=\lambda_k-\mu_k$, we find that the rescue probability can be approximated by
\begin{equation}\label{rescue_prob_approx}
\begin{aligned}
&p_{rescue} \approx \\
  &\begin{cases}
  1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right] ,&
  4\lambda_avp_m>\left(\Delta_a-v\right)^2 ,\\
   1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right] ,&
   \Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 ,\\
   1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] ,&
   \Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 .
  \end{cases}
\end{aligned}
\end{equation}

% TODO do we need this?
%Note that if instead \cref{survprobwapprox1} we apply \cref{survprobwapproxcorrected}, we have
%\begin{align}\label{survprobresistantaneuploidy}
%p_{rescue}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right)\right].
%\end{align}
%However, both approximations seem to work very well (\Cref{compare_1st_2nd_order_approx}).

@@

In this scenario, the probability of evolutionary rescue by cells with aneuploidy from the initial population is
\begin{equation*}
% redefine: p_total, p_denovo=p_resuce from eq above; p_stdvar is the new thing
p_{sgv} = 1-\left(1-p_a\right)^{fN}\approx 1-\e^{-fNp_a}. 
\end{equation*}
The total probability of evolutionary rescue is given by
\begin{align}\nonumber
p_{rescue} 	&= p_{sgv}+\left(1-p_{sgv}\right)p_{de-novo} \\
			&= 1-\exp\left(-\left[\left(1-f\right)p_w + fp_a\right]N\right) .
\end{align}

@@

\subsection*{Effect of aneuploidy on evolutionary rescue}
To determine the extent to which aneuploidy may affect evolutionary rescue, we define $H$ to be the ratio of the rescue probability with aneuploidy ($u>0$) and the rescue probability without aneuploidy ($u=0$),
\begin{equation}\label{ratiorescueexact}
H = \frac{p_{rescue}(u>0)}{p_{rescue}(u=0)}.
\end{equation}

@@

The number of successful mutants generated until time $t$ can be approximated by an inhomogeneous Poisson process with rate $R\left(t\right) = u p_a w_t$,
where $w_t=N\e^{\Delta_w t}$ is the number of wildtype cells at time $t$.
Note that 
\begin{equation}\label{integralR}
\int_0^{t}{R(z)\d z} = 
u p_a N \frac{\exp[{\Delta_w t}]-1}{\Delta_w} \approx 
u p_a N t,
\end{equation}
by integrating the exponential and because $\frac{\exp[\Delta_w t]-1}{\Delta_w}=\frac{1+\Delta_w t+O(t^2)-1}{\Delta_w}=t+O(t^2)$.
The probability density function of $T_1$ is thus
$R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)$. % TODO ref
Therefore, the probability density function of $(T_1 \mid T_1 < \infty)$ is
$f_1(t) = \frac{R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)}{p_{rescue}}$. 
\\

We are interested in the mean conditional time, $\tau_1=\mathbb{E}\left[T_1 \mid T_1<\infty\right]$, which is given by
\begin{equation}\label{meantime1}
\begin{aligned}
\tau_1 =
\int_{0}^{\infty}{t f_1(t) \d t} = 
\frac{\int_{0}^{\infty}{tR(t)\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}} = 
\frac{\int_{0}^{\infty}{\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}}
\end{aligned}
\end{equation}
after applying integration by parts.
Therefore, plugging \cref{integralR,rescue_prob2} in \cref{meantime1}, the probability density function of $T_1$ is given by:
\begin{align}
f_{T_1}\left(t_1\right)=R\left(t_1\right)\e^{-\int_0^{t_1} R(t)\d t}.
\end{align}
As a result, the time $T_1$ conditional on evolutionary rescue is given by:
\begin{align}
f_{T_1}\left(t_1|m_{t\rightarrow\infty}\neq0\right)=\frac{R\left(t_1\right)\e^{-\int_0^{t_1} R(t)\d t}}{1-\left(1-p_w\right)^N}.
\end{align}
The expectation of $T_1$ is:
\begin{align}\label{meantime1}
\tau_1=\mathbb{E}\left[T_1\right]&=\frac{\int_0^\infty\e^{-\int_0^\tau R(t)\d t}\d\tau}{1-\left(1-p_w\right)^N}=\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}{1-\left(1-p_w\right)^N}.
\end{align}
The fration in the exponential of the integrand in \eqref{meantime1} can approximated as:
\begin{equation*}
\frac{\e^{\Delta_w\tau}-1}{\Delta_w}=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2).
\end{equation*}
As a result, the mean time $\tau_1$ can be simplified:
\begin{align}\label{limitapprox}
\tau_1 =
\frac{\int_{0}^{\infty}{\exp\left(-u p_a N t\right) \d t}}{1-\e^{-Np_w}} \approx
\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-u p_a N t} \d t =
\frac{1+\e^{-Np_w}}{u p_a N} ,
\end{align}
where we use the approximation $\left(1-\e^{-Np_w}\right)^{-1}\approx1+\e^{-Np_w}$ and integrate the exponent.
\Cref{MeanTimeGrowthAneuploidyPlot} show the agreement between this approximating and simulation results for intermediate and large tumor sizes.

@@

\section{Methods}

@@

@@

\begin{table}
\begin{center}
  \begin{tabular}{| l |p{5cm}| c | c | p{3cm} |}
    \hline
     & Name & Value & Units & References \\ \hline
    $N$ & Initial tumor size & $10^7-10^9$ & cells  & \citet{del2009does} \\ \hline
    $\lambda_w$ & Wildtype division rate& 0.14 & 1/days  & \citep{bozic2013evolutionary} \\ \hline
    $\mu_w$ & Wildtype death rate& 0.17 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
    $\lambda_a$  & Aneuploid division rate$^\ast$ & 0.14 & 1/days  & - \\ \hline
    $\mu_a$ & Aneuploid death rate$^\ast$ & $0.13-0.17$ & 1/days  & - \\ \hline
    $\lambda_m$ & Mutant division rate& 0.14 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
    $\mu_m$ & Mutant death rate& 0.13 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
    $u$ & Missegregation rate& $10^{-3}-10^{-2}$ & 1$\slash$cell division  & \citet{nowak2004evolutionary,bakker2023predicting} \\ \hline
    $v$ & Mutation rate& $10^{-7}-10^{-9}$ &  1$\slash$gene$\slash$cell division  & \citet{nowak2004evolutionary} \\ 
    % TODO citing Nowak 2004 for the rates is strange because it is a theory paper - cite the original empirical papers
    \hline
  \end{tabular}
\caption{\textbf{Model parameters.} 
Aneuploid birth rate $\lambda_a$ is set to the same value as the wildtype and mutant birth rates, $\lambda_w$ and $\lambda_m$.
Aneuploid death rate $\mu_a$ is set to an intermediate value between the wildtype and mutant death rates, $\mu_w$ and $\mu_m$.}
  \label{table1}
\end{center}
\end{table}

@@

\begin{figure}
% A partial resistance
% B tolerance
% left function of wildtype growth rate 
% right function of population size
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/RatioEvolRescue.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/RatioEvolRescuePopulationSize.pdf}
\end{subfigure}
\caption{\textbf{Effect of aneuploidy on evolutionary rescue.}
The ratio of rescue probability with and without aneuploid ($H$, \cref{ratiorescue}) increases with the aneuploid growth rate (colors) and decreases with the wildtype growth rates and initial tumor size (x-axes), except for large tumors where where the ratio converges to unity.
\textbf{(A-left,A-right)} Aneuploidy provides partial resistance.
\textbf{(B-left, B-right)} Aneuploidy provides tolerance.  
Solid and dashed lines apply $p_{rescue}$ from the exact formula of  (\cref{survival_prob} in \cref{rescue_prob}); dashed lines apply $p_{rescue}$ from the approximate formula (\cref{rescue_prob_approx}), with good agreement.
Parameters: $N=10^4$; $\lambda_w=\lambda_a=\lambda_m=0.14$; (B) $\mu_w=0.17$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
% TODO start y at 1
% TODO fix panel labels
}
\label{rescue_ratio}
\end{figure}

@@

%p_{rescue}=1-\e^{-\left(\theta_1+\theta_2\right)},

>>>>

To estimate the rescue probability $\presc$, we assume independence between clonal lineages starting from an initial population of $N$ wildtype cells (we check the effect of density-dependent growth on our results below).
Thus, the rescue probability is given by 
\begin{subequations}
\begin{align}
\presc &= 1-\left(1-p_w\right)^N \label{rescue_prob} \\
& \approx 1-\e^{-Np_w} , \label{rescue_prob2}
\end{align}
\end{subequations}
where $p_w$ is the probability that the lineage of a single wild-type cell avoids extinction, and the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not $N p_w$) is small.

In the Appendix below, we use the theory of multi-type branching processes to find approximate expressions \cref{survprobwapprox1,survprobwinitial,scenario3} for $p_w$ in different regimes. Substituting these  into \cref{rescue_prob2}, we find that the rescue probability can be approximated by
\begin{equation}\label{rescue_prob_approx}
\begin{aligned}
&\presc \approx \\
  &\begin{cases}
  1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right] ,&
  4\lambda_avp_m>\left(\Delta_a-v\right)^2 ,\\
   1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right] ,&
   \Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 ,\\
   1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] ,&
   \Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 .
  \end{cases}
\end{aligned}
\end{equation}

\daniel{Need to explain intuitively what all this means.}

% TODO do we need this?
%Note that if instead \cref{survprobwapprox1} we apply \cref{survprobwapproxcorrected}, we have
%\begin{align}\label{survprobresistantaneuploidy}
%\presc=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right)\right].
%\end{align}
%However, both approximations seem to work very well (\Cref{compare_1st_2nd_order_approx}).

@@

In this scenario, the probability of evolutionary rescue by cells with aneuploidy from the initial population is
\begin{equation*}
% redefine: p_total, p_denovo=p_resuce from eq above; p_stdvar is the new thing
p_{sgv} = 1-\left(1-p_a\right)^{fN}\approx 1-\e^{-fNp_a}. 
\end{equation*}
The total probability of evolutionary rescue is given by
\begin{align}\nonumber
\presc 	&= p_{sgv}+\left(1-p_{sgv}\right)p_{de-novo} \\
			&= 1-\exp\left(-\left[\left(1-f\right)p_w + fp_a\right]N\right) .
\end{align}

@@

\subsection*{Effect of aneuploidy on evolutionary rescue}
To determine the extent to which aneuploidy may affect evolutionary rescue, we define $H$ to be the ratio of the rescue probability with aneuploidy ($u>0$) and the rescue probability without aneuploidy ($u=0$),
\begin{equation}\label{ratiorescueexact}
H = \frac{\presc(u>0)}{\presc(u=0)}.
\end{equation}

@@

The number of successful mutants generated until time $t$ can be approximated by an inhomogeneous Poisson process with rate $R\left(t\right) = u p_a w_t$,
where $w_t=N\e^{\Delta_w t}$ is the number of wildtype cells at time $t$.
Note that 
\begin{equation}\label{integralR}
\int_0^{t}{R(z)\d z} = 
u p_a N \frac{\exp[{\Delta_w t}]-1}{\Delta_w} \approx 
u p_a N t,
\end{equation}
by integrating the exponential and because $\frac{\exp[\Delta_w t]-1}{\Delta_w}=\frac{1+\Delta_w t+O(t^2)-1}{\Delta_w}=t+O(t^2)$.
The probability density function of $T_1$ is thus
$R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)$. % TODO ref
Therefore, the probability density function of $(T_1 \mid T_1 < \infty)$ is
$f_1(t) = \frac{R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)}{\presc}$. 
\\

We are interested in the mean conditional time, $\tau_1=\mathbb{E}\left[T_1 \mid T_1<\infty\right]$, which is given by
\begin{equation}\label{meantime1}
\begin{aligned}
\tau_1 =
\int_{0}^{\infty}{t f_1(t) \d t} = 
\frac{\int_{0}^{\infty}{tR(t)\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{\presc} = 
\frac{\int_{0}^{\infty}{\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{\presc}
\end{aligned}
\end{equation}
after applying integration by parts.
Therefore, plugging \cref{integralR,rescue_prob2} in \cref{meantime1}, the probability density function of $T_1$ is given by:
\begin{align}
f_{T_1}\left(t_1\right)=R\left(t_1\right)\e^{-\int_0^{t_1} R(t)\d t}.
\end{align}
As a result, the time $T_1$ conditional on evolutionary rescue is given by:
\begin{align}
f_{T_1}\left(t_1|m_{t\rightarrow\infty}\neq0\right)=\frac{R\left(t_1\right)\e^{-\int_0^{t_1} R(t)\d t}}{1-\left(1-p_w\right)^N}.
\end{align}
The expectation of $T_1$ is:
\begin{align}\label{meantime1}
\tau_1=\mathbb{E}\left[T_1\right]&=\frac{\int_0^\infty\e^{-\int_0^\tau R(t)\d t}\d\tau}{1-\left(1-p_w\right)^N}=\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}{1-\left(1-p_w\right)^N}.
\end{align}
The fration in the exponential of the integrand in \eqref{meantime1} can approximated as:
\begin{equation*}
\frac{\e^{\Delta_w\tau}-1}{\Delta_w}=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2).
\end{equation*}
As a result, the mean time $\tau_1$ can be simplified:
\begin{align}\label{limitapprox}
\tau_1 =
\frac{\int_{0}^{\infty}{\exp\left(-u p_a N t\right) \d t}}{1-\e^{-Np_w}} \approx
\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-u p_a N t} \d t =
\frac{1+\e^{-Np_w}}{u p_a N} ,
\end{align}
where we use the approximation $\left(1-\e^{-Np_w}\right)^{-1}\approx1+\e^{-Np_w}$ and integrate the exponent.
\Cref{MeanTimeGrowthAneuploidyPlot} show the agreement between this approximating and simulation results for intermediate and large tumor sizes.

@@

\section{Appendices}

@@

\daniel{Rewrite the below in terms of $\Delta k$.}

@@

\begin{table}
\begin{center}
  \begin{tabular}{| l |p{5cm}| c | c | p{3cm} |}
    \hline
     & Name & Value & Units & References \\ \hline
    $N$ & Initial tumor size & $10^7-10^9$ & cells  & \citet{del2009does} \\ \hline
    $\lambda_w$ & Wildtype division rate& 0.14 & 1/days  & \citep{bozic2013evolutionary} \\ \hline
    $\mu_w$ & Wildtype death rate& 0.17 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
    $\lambda_a$  & Aneuploid division rate$^\ast$ & 0.14 & 1/days  & - \\ \hline
    $\mu_a$ & Aneuploid death rate$^\ast$ & $0.13-0.17$ & 1/days  & - \\ \hline
    $\lambda_m$ & Mutant division rate& 0.14 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
    $\mu_m$ & Mutant death rate& 0.13 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
    $\Delta_k \equiv \lambda_k - \mu_k$ & 1/days & Growth rate of type $k$ cells, for $k=w,a,m$\\ \hline
    $u$ & Missegregation rate& $10^{-3}-10^{-2}$ & 1$\slash$cell division  & \citet{nowak2004evolutionary,bakker2023predicting} \\ \hline
    $v$ & Mutation rate& $10^{-7}-10^{-9}$ &  1$\slash$gene$\slash$cell division  & \citet{nowak2004evolutionary} \\ 
    % TODO citing Nowak 2004 for the rates is strange because it is a theory paper - cite the original empirical papers
    \hline
  \end{tabular}
\caption{\textbf{Model parameters.} 
Aneuploid birth rate $\lambda_a$ is set to the same value as the wildtype and mutant birth rates, $\lambda_w$ and $\lambda_m$.
Aneuploid death rate $\mu_a$ is set to an intermediate value between the wildtype and mutant death rates, $\mu_w$ and $\mu_m$.}
  \label{table1}
\end{center}
\end{table}

@@

\begin{figure}
% A partial resistance
% B tolerance
% left function of wildtype growth rate 
% right function of population size
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/RatioEvolRescue.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/RatioEvolRescuePopulationSize.pdf}
\end{subfigure}
\caption{\textbf{Effect of aneuploidy on evolutionary rescue.}
The ratio of rescue probability with and without aneuploid ($H$, \cref{ratiorescue}) increases with the aneuploid growth rate (colors) and decreases with the wildtype growth rates and initial tumor size (x-axes), except for large tumors where where the ratio converges to unity.
\textbf{(A-left,A-right)} Aneuploidy provides partial resistance.
\textbf{(B-left, B-right)} Aneuploidy provides tolerance.  
Solid and dashed lines apply $\presc$ from the exact formula of  (\cref{survival_prob} in \cref{rescue_prob}); dashed lines apply $\presc$ from the approximate formula (\cref{rescue_prob_approx}), with good agreement.
Parameters: $N=10^4$; $\lambda_w=\lambda_a=\lambda_m=0.14$; (B) $\mu_w=0.17$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
% TODO start y at 1
% TODO fix panel labels
}
\label{rescue_ratio}
\end{figure}

@@

%\presc=1-\e^{-\left(\theta_1+\theta_2\right)},
